\textbf{How open are hybrid journals included in transformative agreements?}

### 2024-02-19 analysis/manuscript.Rmd
starting discussions

For over two decades, hybrid open access journal publishing, which makes some articles openly available while others remain behind a paywall, has been discussed as a means for transitioning the subscription system to full open access [@Prosser_2003].
The idea was that when journals increasingly publish open access articles, they could reduce revenues from subscriptions, while libraries and funders could change their funding models and shift expenditures from subscription to open access. 
However, initial approaches, mainly based on publication fees, also called article processing charges (APCs), did not contribute substantially to a large open access uptake. 
In 2009, Springer reported open access to 1% of articles in hybrid journals [@Dallmeier_2010].
Other publishers also recorded a low uptake.
In 2011, only 1-2% of articles were open access [@Bj_rk_2012], growing to around 4% between 2011 and 2013 [@Laakso_2016].

With the introduction of central funding mechanisms for hybrid open access publication fees in some European countries since 2012, an increase in open access could be observed [@Bj_rk_2017; @Piwowar_2018; @Jubb_2017; @Huang_2020].
For example, studying university output, @Robinson_Garcia_2020 estimated a median uptake of 7.1% in the period 2014-2017. 
In particular, British (17%), Austrian (15%) and Dutch (13%) universities contributed to this trend. 
However, this shift in funding policy towards hybrid open access also added to the overall cost of publishing, which includes subscription spending and the administrative efforts required to handle payments [@Pinfield_2016].
Moreover, established large commercial publishers, which already dominated the publishing market [@Larivi_re_2015], disproportionately benefited from hybrid open access funding in comparison to full open access publishers [@Jahn_2016; @Butler_2023; @Shu_2023].

As a consequence, libraries and their consortia began to develop licensing strategies aimed at avoiding such ‘double dipping’ scenarios, in which well-established commercial publishers gain twice from reading and publication fees, as well as to increase publisher-provided immediate open access [@BJ_RK_2014; @Schimmer_2015].
These considerations resulted into transformative agreements, which cover a broad range of successfully negotiated contracts between library consortia and publishers from mid-2010s onwards where institutional spending for subscriptions are re-purposed for open access publishing [@Borrego_2020; @Hinchliffe_2019].
Transformative agreements seek to control costs while allowing a transitional phase for publishing more open access articles. 
Although transformative agreements mainly bundle hybrid and subscription-only journals from commercial publishers, they aim at a higher degree of transparency than previous big deals, where contracts including payments were confidential [@Bergstrom_2014].
<!-- For example, the nationwide DEAL consortium in Germany, which includes universities and academic institutions, collectively cancelled their subscriptions with Elsevier from 2017 onwards to negotiate a transformative agreement [@Fraser_2023]. -->

The introduction of transformative agreements aligns with funding policy changes, such as the decision made by the cOAlition S, a consortium of national funders including the European Commission, to no longer provide financial support for individual publication fees when publishing in hybrid journals.
According ot its Plan S launched in 2018, the cOAlition S funders only accept hybrid open access through transformative agreements "during a transition period that should be as short as possible" [@Schiltz_2018]. 
Specifically, they agreed to support hybrid open access solely through transformative agreements from 1 January 2021, until the end of 2024.
Notably, the German Research Foundation (DFG), despite not being part of cOAlition S, has also extended its financial support for hybrid open access through transformative agreements [@Mittermaier_2021]. Previously, the DFG only provided funding for full open access journals [@Jahn_2016].

By the end of 2023, many transformative agreements were implemented, but outcomes were mixed.
The ESAC Transformative Agreement Registry[^esacregistry], the largest source of disclosure, recorded more than 800 transformative agreements, resulting in up to 900.000 open access articles published in both full open access and hybrid journals according to the accompanying ESAC Market Watch[^esacmarketwatch]. 
Library consortia reported increased open access volume, streamlined payment and monitoring procedures, as well as extensive utilization of open access options by the researchers they serve [@Marques_2020; @Pinhasi_2020; @Parmhed_2023]. 
The ongoing standardisation of transformative agreements contributed to improved transparency in terms of contracts and publisher-provided article metadata [@Marques_2019].
However, with the growing trend toward transformative agreements, continued reliance on big deals is perceived as problematic, which might perpetuate market concentration [@Butler_2023; @Shu_2023]. 
Whether transformative agreements lead to reduced pricing remains uncertain [@Borrego_2023] and
a substantial transition of hybrid journals towards full open access could not be observed [@Momeni_2021; @Matthias_2019].
Furthermore, the focus on large commercial publishers might increase inequality, because transformative agreements continue to target institutions from high-income countries [@Klebel_2023]. 
Besides, an editorial-board resignation raised concerns that publishers' desire to maximize journal publication volume "without regard to quality" is a consequence of transformative agreements [@Goodin_2023].

The controversies surrounding hybrid open access and transformative agreements have led to varying policy conclusions. 
For instance, the  Association of Swedish Higher Education Institutions (Sveriges universitets- och
högskoleförbund, SUHF) recommended to only support agreements for publishing in full open access journals [^sweden].
Likewise, most cOAlition S funders will discontinue financial support for transformative agreements from 2024 onwards [@Liverpool_2023].
The consortium also removed most hybrid journals from its Transformative Journal program in 2023 due to their failure to meet self-defined open access growth targets [@Brainard_2023]. 
In contrast, the German DEAL consortium signed a five-year transformative agreement with Elsevier starting in 2024, while also renewing its contracts with Springer Nature and Wiley until the end of 2028.

Despite these controversies around transformative agreements as a means of transitioning journal publishing to full open access, there is limited evidence available on the adoption of open access in hybrid journals, and the extent to which this uptake can be attributed to transformative agreements.
Previous studies have focused on specific countries [@P_l_nen_2020; @Huang_2020; @Haucap_2021; @Wenaas_2022; @Taubert_2023] or publisher portfolios [@Pieper_2018; @Bakker_2024; @Fraser_2023; @Jahn_2021; @Momeni_2023], while related large-scale studies relied on self-reported agreement data [@Moskovkin_2022], or used APC pricing lists [@Shu_2023]. 
Particularly, data availability is a limiting factor when studying the impact of transformative agreements [@Bakker_2024], because bibliometric databases, even though many allow the retrieval of open access articles in hybrid journals, do not directly attribute them to specific transformative agreements.

The present study aims to address these limitations by combining multiple openly available data sources to determine the open access uptake in hybrid journals, while distinguishing between open access through transformative agreements and other means. 
With this novel and open approach, this study aims to answer the following questions:

- What was the number and proportion of open access articles in hybrid journals included in transformative agreements between 2018 and 2022?
- How did transformative agreements impact open access uptake in hybrid journals?

@@

# Methods and data

This study combines data from multiple publicly available data sources as diagrammatically shown in Figure \ref{fig:data_workflow}. 
Initially, transformative agreement data retrieved from the cOAlition S Journal Checker Tool provided information about journal portfolios and participating institutions. 
After identification of hybrid journals by excluding full open access journals, Crossref served as the primary data source for article-level metadata including Creative Commons (CC) license information to indicate open access availability on publisher websites.
To determine open access articles published through transformative agreements, first author affiliations from OpenAlex [@priem2022openalex] were subsequently linked to eligible institutions according to the transformative agreement data.

@@

Because transformative agreements can include both full open access and hybrid journals, the data were complemented with information about a journal's open access status using multiple sources: the Directory of Open Access Journals (DOAJ) downloaded on 12 December 2023[^doaj], OpenAlex (November 2023) and the the Bielefeld list of GOLD OA journals [@goldoa]. 
As shown in Figure \ref{fig:method_fig}A, combining different data sources considerably extended the journal matching to exclude full open access journals from transformative agreements.
In total, `r n_oa_journals` full open access journals were excluded based on ISSN matching.
The overlap between the three data sources was `r round(length(unique(Reduce(intersect, my_venn))) / length(unique(Reduce(union, my_venn)))* 100,0)`%.
The Gold OA journals dataset alone added `r length(unique(Reduce(setdiff, my_venn)))` journals, while the DOAJ comprised `r length(setdiff(my_venn$DOAJ, union(my_venn[["GOLD OA"]], my_venn$OpenAlex)))` full open access journals not listed in either of the other two sources. 
These full open access journals were mostly launched in 2022.

@@

Open access articles in hybrid journals were identified through Creative Commons (CC) license information in Crossref metadata. 
License information relative to the "accepted manuscript (AM)" version were not considered. 
Crossref was used for open access identification because transformative agreements workflows generally require publishers to deliver CC license information to this DOI registration agency [@Geschuhn_2017].
Comparing Crossref license coverage with OpenAlex, which re-uses open access evidence from Unpaywall, a widely used open access discovery service that also parses journal websites for open content licenses [@Piwowar_2018], highlighted ongoing challenges to identify hybrid open access [@Mart_n_Mart_n_2018; @Jahn_2021; @Zhang_2022; @Butler_2023].
Here, `r hoaddata::cr_upw |> filter(cr_year %in% 2018:2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, cr_hybrid_total) |> pull(cr_hybrid_total) |> sum()` articles with CC license were retrieved using Crossref, while `r hoaddata::cr_upw |> filter(cr_year %in% 2018:2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, upw_hybrid_total) |> pull(upw_hybrid_total) |> sum()` articles were tagged as "hybrid" according to the OpenAlex November 2023 release, which was used throughout this study.
The biggest differences concerned articles published between 2018 and 2020.
In 2022, however, Crossref and OpenAlex open access numbers only differ slightly (`r hoaddata::cr_upw |> filter(cr_year %in% 2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, cr_hybrid_total) |> pull(cr_hybrid_total) |> sum()` records using Crossref vs. `r hoaddata::cr_upw |> filter(cr_year %in% 2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, upw_hybrid_total) |> pull(upw_hybrid_total) |> sum()` in OpenAlex).
Notable difference could be furthermore observed among some publishers that presumably did not provide CC license information to Crossref including AIP Publishing, American Physiological Society and Emerald.
Crossref license metadata was more complete with regard to articles from the publisher Wiley and American Chemical Society.
Finally, inconsistent open access status information in previous OpenAlex versions was observed [@jahn2023]. According to the OpenAlex release notes, fixing this issue was still ongoing, which might also explain this discrepancy.

@@

Affiliation metadata about corresponding authors are crucial for the planning and evaluation of transformative agreements, because they are considered to be responsible to arrange open access publication [@Schimmer_2015; @Geschuhn_2017; @Borrego_2020]. 
Here, country and institutional affiliations were retrieved from OpenAlex.
However, because of low coverage in OpenAlex, this study focused on first authors and their affiliations instead.
First authors typically contribute most to a paper and are often considered lead author research papers [@Larivi_re_2016], and can be therefore assumed as a proxy to measure to open access payments [@Zhang_2022; @Shu_2023].
Overall, around 90% of studied articles had first author affiliation metadata in OpenAlex, whereas the coverage of articles with corresponding author information was around 54%.

@@

To assess the impact of transformative agreements to hybrid open access,
participating institutions from the Transformative Agreement Data dump,
which were crowd-sourced from the agreements and consortia that
successfully negotiated an agreement, were matched with first author
affiliations recorded by OpenAlex using the ROR-ID.
The matching also took into account the duration of an agreement according to the ESAC registry.
Upon inspection, Transformative Agreement Data did not cover associated institutions comprehensively like university hospitals or institutes of large research organisations like the Max Planck Society. To improve the matching, Transformative Agreement Data was automatically enriched with associated organisations using OpenAlex's institution entity.

In total, the compiled data set consists of `r hoaddata::jn_ind |> filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |> distinct(issn_l, cr_year, jn_all) |> pull(jn_all) |> sum()` articles published in `r hoaddata::jn_ind |> filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |> distinct(issn_l) |> nrow()` hybrid journals between 2018 and 2022 (see Figure \ref{fig:method_fig}B). 
Hybrid journals included in transformative agreements represented `r cr_stats |> filter(src == "hoad") |> pull(n_articles) |> sum() / cr_stats |> pull(n_articles) |> sum() * 100`%  of total global output over the same time period according to Crossref, while full open access journals accounted for
`r cr_stats |> filter(is.na(src) & oa_status == "gold") |> pull(n_articles) |> sum() / cr_stats |> pull(n_articles) |> sum() * 100`% of total article volume.

@@

Throughout this mostly automated data gathering and analysis process, tools from the Tidyverse (Wickham et al., 2019) for the R programming language (R Core Team, 2020) were used. 
The resulting data is openly available through an R package, hoaddata, version 0.2.91 
Following @Marwick_2018, hoaddata contains not only the datasets used in the data analysis. 
It also includes code used to compile the data by connecting it the cloud-based Google Big Query data warehouse, where the big scholarly data from Crossref, OpenAlex and Unpaywall were imported.
To increase the computational reproducibility, the R package was automatically built using GitHub Actions, a continuous integration service.

@@

Figure \ref{fig:results_overview}B highlights that the majority of open access articles in hybrid journals were made available through transformative agreements in 2021 and 2022, contributing `r oa_ta_plot_df |> filter(cr_year == "2022", type == "ta_oa") |> pull(value) / oa_by_year |> filter(cr_year == "2022") |> pull(oa) * 100`% of the total open access article volume in 2022.
However, there was also a notable growth in open access provision through individual publication fees, which increased from `r oa_ta_plot_df |> filter(cr_year == "2018", type == "no_ta") |> mutate(prop = value / n * 100) |> pull(prop)`% (n = `r oa_ta_plot_df |> filter(cr_year == "2018", type == "no_ta") |> pull(value)`) in 2018 to  `r oa_ta_plot_df |> filter(cr_year == "2022", type == "no_ta") |> mutate(prop = value / n * 100) |> pull(prop)`% (n = `r oa_ta_plot_df |> filter(cr_year == "2022", type == "no_ta") |> pull(value)`). This suggests that publishers were able to gain equally from individual and institutional open access publishing options.

Figure \ref{fig:results_overview}C depicts the substantial variations among the hybrid journals included in transformative agreements in terms of open access uptake.
Although the median generally follows the trend shown in Figure \ref{fig:results_overview}A, the farther stretch of upper quartiles and whiskers over the years illustrates that an increasing number of journals published an above-average proportion of open access articles. 
In 2022, 25% of hybrid journals (n = `r oa_jn_box_df |> filter(cr_year == "2022") |> mutate(top25 = quantile(prop, .75)) |> filter(prop >= top25) |> nrow()`) had an open access uptake of `r oa_jn_box_df |> filter(cr_year == "2022") |> pull(prop) |> quantile(0.75)*100`%, and `r (oa_jn_box_df |> filter(cr_year == "2022", prop >= 0.5) |> nrow()) / length(unique(oa_jn_box_df$issn_l)) *100`% of journals (n = `r oa_jn_box_df |> filter(cr_year == "2022", prop >= 0.5) |> nrow()`) provided the majority of their articles under a Creative Commons license in the same year. 
These journals were, on average, smaller (M = `r oa_jn_box_df |> filter(cr_year == "2022", prop >= 0.5) |> pull(n) |> mean()`, SD = `r oa_jn_box_df |> filter(cr_year == "2022", prop >= 0.5) |> pull(n) |> sd()`) than those with an open access share below 50% (M = `r oa_jn_box_df |> filter(cr_year == "2022", prop <= 0.5) |> pull(n) |> mean()`, SD = `r oa_jn_box_df |> filter(cr_year == "2022", prop <= 0.5) |> pull(n) |> sd()`). 
Notable exception of large journals with an above-average open access proportion were Physical Review D, a high-energy physics journal covered by the SCOAP3 consortium that provided open access to `r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "2470-0010") |> pull(oa_all)` out of `r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "2470-0010") |> pull(n)` articles in 2022, Astronomy and Physics (`r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "0004-6361") |> pull(oa_all)` out of `r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "0004-6361") |> pull(n)` articles in 2022 were open access), which shifted to a subscribe to open business model for all accepted articles as of April 2022, and the Journal of Fluid Mechanics (`r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "0022-1120") |> pull(oa_all)` of `r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "0022-1120") |> pull(n)` articles in 2022 were open access).

When comparing the impact of open access trough transformative agreements across journals, it shows that for many journals these agreements substantially contributed to the growth of open access over the years (Figure \ref{fig:results_overview}D). 
Examples of such journals include those with a scope on specific countries or regions, where also transformative agreements were implemented. 
For instance, in 2022, the Germany-based journals Zeitschrift für Erziehungswissenschaft and Zeitschrift für Politikwissenschaft, as well as the Scandinavian Political Studies adressing the Nordic countries, achieved an overall open access uptake of more than 90% just through transformative agreements. 
Despite the rise in transformative agreements, it is worth noting that other means of publishing open access in hybrid journals remained common. 
In total, `r summary_all |> filter(cr_year == "2022", value > 0, name == "prop_no_ta") |> nrow()` journals published open access articles from authors affiliated with institutions without transformative agreements in place, while `r summary_all |> filter(cr_year == "2022", value > 0, name == "prop_ta_oa") |> nrow()` journals published at least one open access article through a transformative agreement in the same year.

@@

>>>>

For over two decades, hybrid open access journal publishing, which makes some articles openly available while others remain behind a paywall, has been discussed as a means for transitioning the subscription system to full open access [@Prosser_2003].
The idea was that when journals increasingly publish open access articles, they could reduce revenues from subscriptions, while libraries and funders could change their funding models and shift expenditures from subscription to open access. 
However, initial approaches, mainly based on publication fees, also called article processing charges (APCs), did not contribute substantially to a large open access uptake. 
In 2009, Springer reported open access to 1% of articles in hybrid journals [@Dallmeier_2010].
Other studies also recorded a low uptake.
In 2011, only 1-2% of articles were open access [@Bj_rk_2012], growing to around 4% between 2011 and 2013 [@Laakso_2016].

With the introduction of central funding mechanisms for publication fees in some European countries since 2012, an increase in hybrid open access could be observed [@Bj_rk_2017; @Piwowar_2018; @Jubb_2017; @Huang_2020].
For example, studying university output, @Robinson_Garcia_2020 estimated a median uptake of 7.1% in the period 2014-2017. 
In particular, British (17%), Austrian (15%) and Dutch (13%) universities contributed to this trend. 
However, this shift in funding policy towards hybrid open access also added to the overall cost of publishing, which includes subscription spending and the administrative efforts required to handle payments [@Pinfield_2016].
Moreover, established large commercial publishers, which already dominated the publishing market [@Larivi_re_2015], disproportionately benefited from hybrid open access funding in comparison to full open access publishers [@Jahn_2016; @Butler_2023; @Shu_2023].

As a consequence, libraries and their consortia began to develop licensing strategies aimed at avoiding such ‘double dipping’ scenarios, in which well-established commercial publishers gain twice from reading and open access publishing, as well as to increase publisher-provided immediate open access [@BJ_RK_2014; @Schimmer_2015].
These considerations resulted into transformative agreements, which cover a broad range of contracts between library consortia and publishers from mid-2010s onwards where institutional spending for subscriptions and open access publishing are considered together [@Borrego_2020; @Hinchliffe_2019].
Transformative agreements seek to control costs while allowing a transitional phase for publishing more open access articles. 
Similar to big deals, transformative agreements mainly bundle hybrid and subscription-only journals from commercial publishers, but aim at a higher degree of transparency than previous big deals, where contracts including payments were confidential [@Bergstrom_2014].
<!-- For example, the nationwide DEAL consortium in Germany, which includes universities and academic institutions, collectively cancelled their subscriptions with Elsevier from 2017 onwards to negotiate a transformative agreement [@Fraser_2023]. -->

The introduction of transformative agreements aligns with funding policy changes, such as the decision made by the cOAlition S, a consortium of national funders including the European Commission, to no longer provide financial support for individual publication fees when publishing in hybrid journals.
According ot its Plan S launched in 2018, the cOAlition S funders only accept hybrid open access through transformative agreements "during a transition period that should be as short as possible" [@Schiltz_2018]. 
Specifically, they agreed to support hybrid open access only through transformative agreements from 2021, until the end of 2024.
Notably, the German Research Foundation (DFG), despite not being part of cOAlition S, has also extended its financial support for hybrid open access through transformative agreements [@Mittermaier_2021]. Previously, the DFG only provided funding for full open access journals [@Jahn_2016].

By the end of 2023, many transformative agreements were implemented, but interim outcomes are mixed.
The ESAC Transformative Agreement Registry[^esacregistry], the largest source of disclosure, recorded more than 800 transformative agreements, resulting in up to 900.000 open access articles published in both full open access and hybrid journals according to the accompanying ESAC Market Watch[^esacmarketwatch]. 
Library consortia reported increased open access volume, streamlined payment and monitoring procedures, as well as extensive utilization of open access options by the researchers they serve [@Marques_2020; @Pinhasi_2020; @Parmhed_2023]. 
The ongoing standardisation of transformative agreements contributed to improved transparency in terms of contracts and publisher-provided article metadata [@Marques_2019; @Pinhasi_2021].
However, with the growing trend toward transformative agreements, continued reliance on big deals is perceived as problematic, because it perpetuate market concentration [@Butler_2023; @Shu_2023]. 
Whether transformative agreements lead to reduced pricing remains uncertain [@Borrego_2023] and
a substantial transition of hybrid journals towards full open access could not be observed [@Momeni_2021; @Matthias_2019].
The focus on large commercial publishers might also increase inequality, because transformative agreements focus on pay to publish open access mainly targets institutions from high-income countries, furthering a questionable journal prestige culture [@budapest]. 
Besides, an editorial-board resignation raised concerns that publishers' desire to maximize journal publication volume "without regard to quality" is a consequence of transformative agreements [@Goodin_2023].

The controversies surrounding hybrid open access and transformative agreements have led to varying policy conclusions. 
For instance, the Association of Swedish Higher Education Institutions (Sveriges universitets- och
högskoleförbund, SUHF) recommended to only support agreements for publishing in full open access journals [^sweden].
Likewise, most cOAlition S funders will discontinue financial support for transformative agreements by the end of 2024 [@Liverpool_2023].
The consortium also removed most hybrid journals from its Transformative Journal program in 2023 due to publishers' failure to meet self-defined open access growth targets [@Brainard_2023]. 
In contrast, the German DEAL consortium signed a five-year transformative agreement with Elsevier starting in 2024, while also renewing its contracts with Springer Nature and Wiley until the end of 2028.

Despite these controversies around transformative agreements as a means of transitioning journal publishing to full open access, there is limited evidence available on the adoption of open access in hybrid journals, and the extent to which this uptake can be attributed to transformative agreements.
Previous studies have focused on specific countries [@P_l_nen_2020; @Huang_2020; @Haucap_2021; @Wenaas_2022; @Taubert_2023] or publisher portfolios [@Pieper_2018; @Bakker_2024; @Fraser_2023; @Jahn_2021; @Momeni_2023], while large-scale studies relied on self-reported agreement data [@Moskovkin_2022], or used APC pricing lists [@Shu_2023]. 
Particularly, data availability is a limiting factor when studying the impact of transformative agreements [@Bakker_2024], because bibliometric databases, even though many allow the retrieval of open access articles in hybrid journals, do not directly attribute them to specific transformative agreements.

The present study aims to address these limitations by combining multiple openly available data sources to determine the open access uptake in hybrid journals, while distinguishing between open access through transformative agreements and other means. 
With this novel and open approach, this the first large-scale analysis aims to answer the following questions:

- What was the number and proportion of open access articles in hybrid journals in transformative agreements between 2018 and 2022?
- To what extent did institutions with a transformation agreement contribute to the adoption of open access in hybrid journals?

@@

# Methods

This study combines data from multiple publicly available data sources as diagrammatically shown in Figure \ref{fig:data_workflow}. 
Initially, transformative agreement data retrieved from the cOAlition S Journal Checker Tool[^checkertool]  provided information about journal portfolios and participating institutions. 
After identification of hybrid journals by excluding full open access journals, Crossref served as the primary data source for article-level metadata including Creative Commons (CC) license information to indicate open access availability on publisher websites.
To determine open access articles published through transformative agreements, first author affiliations from OpenAlex [@priem2022openalex] were subsequently linked to eligible institutions according to the transformative agreement data. 
In the following, these steps are described in more detail.

@@

Because transformative agreements can include both full open access and hybrid journals, the data were complemented with information about a journal's open access status using multiple sources: the Directory of Open Access Journals (DOAJ) downloaded on 12 December 2023[^doaj], OpenAlex (November 2023) and the the Bielefeld list of GOLD OA journals [@goldoa]. 
As shown in Figure \ref{fig:method_fig}A, combining different data sources considerably extended the journal matching<!-- to exclude full open access journals from transformative agreements-->.
In total, `r n_oa_journals` full open access journals were excluded based on ISSN matching.
The overlap between the three data sources was `r round(length(unique(Reduce(intersect, my_venn))) / length(unique(Reduce(union, my_venn)))* 100,0)`%.
The Gold OA journals dataset alone added `r length(unique(Reduce(setdiff, my_venn)))` journals, while the DOAJ comprised `r length(setdiff(my_venn$DOAJ, union(my_venn[["GOLD OA"]], my_venn$OpenAlex)))` full open access journals not listed in either of the other two sources. 
These full open access journals were mostly launched in 2022.

@@

Open access articles in hybrid journals were identified through Creative Commons (CC) license information in Crossref metadata. 
License information relative to the "accepted manuscript (AM)" version were not considered. 
Crossref was used for open access identification because transformative agreements workflows generally require publishers to deliver CC license information to this DOI registration agency [@Geschuhn_2017].
Comparing Crossref license coverage with OpenAlex, which re-uses open access evidence from Unpaywall, a widely used open access discovery service that also parses journal websites for open content licenses [@Piwowar_2018], highlighted ongoing challenges to identify hybrid open access [@Mart_n_Mart_n_2018; @Jahn_2021; @Zhang_2022; @Butler_2023].
Here, `r hoaddata::cr_upw |> filter(cr_year %in% 2018:2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, cr_hybrid_total) |> pull(cr_hybrid_total) |> sum()` articles with CC license were retrieved using Crossref, while `r hoaddata::cr_upw |> filter(cr_year %in% 2018:2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, upw_hybrid_total) |> pull(upw_hybrid_total) |> sum()` articles were tagged as "hybrid" according to the OpenAlex November 2023 release, which was used throughout this study.
The biggest differences concerned articles published between 2018 and 2020.
In 2022, however, Crossref and OpenAlex open access numbers only differ slightly (`r hoaddata::cr_upw |> filter(cr_year %in% 2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, cr_hybrid_total) |> pull(cr_hybrid_total) |> sum()` records using Crossref vs. `r hoaddata::cr_upw |> filter(cr_year %in% 2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, upw_hybrid_total) |> pull(upw_hybrid_total) |> sum()` in OpenAlex).
Notable difference could be furthermore observed among some publishers that presumably did not provide CC license information to Crossref including AIP Publishing, American Physiological Society, Emerald and the Royal Society.
Crossref license metadata was more complete with regard to articles from the publisher Wiley and American Chemical Society.
Finally, inconsistent open access status information in previous OpenAlex versions was observed [@jahn2023]. After reporting it to OpenAlex, according to the release notes, fixing this issue was still ongoing, which might also explain part of the discrepancy.

@@

Affiliation metadata about corresponding authors are crucial for the planning and evaluation of transformative agreements, because they are considered to be responsible to arrange open access publication [@Schimmer_2015; @Geschuhn_2017; @Borrego_2020]. 
Here, country and institutional affiliations were retrieved from OpenAlex.
However, because of low coverage in OpenAlex, this study focused on first authors and their affiliations instead.
First authors typically contribute most to a paper and are often considered lead author research papers [@Larivi_re_2016]. 
Related studies therefore assumed first authors as a proxy to measure to open access payments and the impact of transformative agreement [@Zhang_2022; @Shu_2023; @Haucap_2021].
Overall, around 90% of studied articles had first author affiliation metadata in OpenAlex, whereas the coverage of articles with corresponding author information was around 54%.

@@

To assess the impact of transformative agreements to hybrid open access,
participating institutions from the Transformative Agreement Data dump,
which were crowd-sourced from the agreements and consortia that
successfully negotiated an agreement, were matched with first author
affiliations recorded by OpenAlex using the ROR-ID.
The matching also took into account the duration of an agreement according to the ESAC registry.
Upon inspection, Transformative Agreement Data did not cover associated institutions comprehensively like university hospitals or institutes of large research organisations like the Max Planck Society. To improve the matching, Transformative Agreement Data was automatically enriched with associated organisations using OpenAlex's institution entity data.

In total, the compiled data set consists of `r hoaddata::jn_ind |> filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |> distinct(issn_l, cr_year, jn_all) |> pull(jn_all) |> sum()` articles published in `r hoaddata::jn_ind |> filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |> distinct(issn_l) |> nrow()` hybrid journals between 2018 and 2022 (see Figure \ref{fig:method_fig}B). 
Hybrid journals in transformative agreements represented `r cr_stats |> filter(src == "hoad") |> pull(n_articles) |> sum() / cr_stats |> pull(n_articles) |> sum() * 100`%  of total global output over the same time period according to Crossref, while full open access journals recorded 
`r cr_stats |> filter(is.na(src) & oa_status == "gold") |> pull(n_articles) |> sum() / cr_stats |> pull(n_articles) |> sum() * 100`%.

@@

Throughout this mostly automated data gathering and analysis process, tools from the Tidyverse (Wickham et al., 2019) for the R programming language (R Core Team, 2020) were used. 
The resulting data is openly available through an R data package, hoaddata, version 0.2.91 
Following @Marwick_2018, hoaddata contains not only the datasets used in the data analysis. 
It also includes code used to compile the data by connecting it the cloud-based Google Big Query data warehouse, where the big scholarly data from Crossref, OpenAlex and Unpaywall were imported.
To increase the computational reproducibility, the data aggregation through hoaddata was automatically carried out using GitHub Actions, a continuous integration service.

@@

Figure \ref{fig:results_overview}B highlights that the majority of hybrid open access was made available through transformative agreements in 2021 and 2022, contributing `r oa_ta_plot_df |> filter(cr_year == "2022", type == "ta_oa") |> pull(value) / oa_by_year |> filter(cr_year == "2022") |> pull(oa) * 100`% of the total open access article volume in 2022.
However, there was also a notable growth in open access provision through other means, most likely publication fees being not invoiced through transformative agreements, which increased from `r oa_ta_plot_df |> filter(cr_year == "2018", type == "no_ta") |> mutate(prop = value / n * 100) |> pull(prop)`% (n = `r oa_ta_plot_df |> filter(cr_year == "2018", type == "no_ta") |> pull(value)`) in 2018 to  `r oa_ta_plot_df |> filter(cr_year == "2022", type == "no_ta") |> mutate(prop = value / n * 100) |> pull(prop)`% (n = `r oa_ta_plot_df |> filter(cr_year == "2022", type == "no_ta") |> pull(value)`). 
<!-- This suggests that publishers were able to gain equally from individual and institutional open access publishing options.-->

Figure \ref{fig:results_overview}C depicts the substantial variations among the hybrid journals included in transformative agreements in terms of open access uptake.
Although the median generally follows the trend shown in Figure \ref{fig:results_overview}A, the farther stretch of upper quartiles and whiskers over the years illustrates that an increasing number of journals published an above-average proportion of open access articles. 
In 2022, 25% of hybrid journals (n = `r oa_jn_box_df |> filter(cr_year == "2022") |> mutate(top25 = quantile(prop, .75)) |> filter(prop >= top25) |> nrow()`) had an open access uptake of `r oa_jn_box_df |> filter(cr_year == "2022") |> pull(prop) |> quantile(0.75)*100`%, and `r (oa_jn_box_df |> filter(cr_year == "2022", prop >= 0.5) |> nrow()) / length(unique(oa_jn_box_df$issn_l)) *100`% of journals (n = `r oa_jn_box_df |> filter(cr_year == "2022", prop >= 0.5) |> nrow()`) provided the majority of their articles under a Creative Commons license in the same year. 
These journals were, on average, smaller (M = `r oa_jn_box_df |> filter(cr_year == "2022", prop >= 0.5) |> pull(n) |> mean()`, SD = `r oa_jn_box_df |> filter(cr_year == "2022", prop >= 0.5) |> pull(n) |> sd()`) than those with an open access share below 50% (M = `r oa_jn_box_df |> filter(cr_year == "2022", prop <= 0.5) |> pull(n) |> mean()`, SD = `r oa_jn_box_df |> filter(cr_year == "2022", prop <= 0.5) |> pull(n) |> sd()`). 
<!-- Notable exception of large journals with an above-average open access proportion were Physical Review D, a high-energy physics journal covered by the SCOAP3 consortium that provided open access to `r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "2470-0010") |> pull(oa_all)` out of `r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "2470-0010") |> pull(n)` articles in 2022, Astronomy and Physics (`r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "0004-6361") |> pull(oa_all)` out of `r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "0004-6361") |> pull(n)` articles in 2022 were open access), which shifted to a subscribe to open business model for all accepted articles as of April 2022, and the Journal of Fluid Mechanics (`r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "0022-1120") |> pull(oa_all)` of `r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "0022-1120") |> pull(n)` articles in 2022 were open access). -->

When comparing the impact of open access trough transformative agreements across journals, it shows that for many journals these agreements substantially contributed to the growth of open access over the years (Figure \ref{fig:results_overview}D). 
<!-- Examples of such journals include those with a scope on specific countries or regions, where also transformative agreements were implemented. 
For instance, in 2022, the Germany-based journals Zeitschrift für Erziehungswissenschaft and Zeitschrift für Politikwissenschaft, as well as the Scandinavian Political Studies adressing the Nordic countries, achieved an overall open access uptake of more than 90% just through transformative agreements. -->
Despite the rise in transformative agreements, it is worth noting that other means of publishing open access remained common across the investiagted hybrid journals. 
In total, `r summary_all |> filter(cr_year == "2022", value > 0, name == "prop_no_ta") |> nrow()` journals published open access articles from authors affiliated with institutions without transformative agreements in place, while `r summary_all |> filter(cr_year == "2022", value > 0, name == "prop_ta_oa") |> nrow()` journals published at least one open access article through a transformative agreement in the same year.

@@

The primary aim of this study was to investigate the adoption of open access in hybrid journals through transformative agreements, presenting a novel approach based on open data. 
The results indicate a steady increase, estimating that the majority of open access in hybrid journals in 2022 originate from institutions that had an transformative agreement in place. 
Yet, the majority of research literature published in hybrid journals between 2018 and 2022 remained behind a publisher paywall. 
Growth in the adoption of open access in hybrid journals through transformative agreements can be largely attributed to big three commercial publishers -- Elsevier, Springer Nature, and Wiley -- but varies across journals, publishers, subjects, and country affiliations. 
Despite limitations regarding the data used, the findings indicate that the current level of implementation of transformative agreements is insufficient to bring about a large-scale transformation to full open access.
